---
title: "Data Documentation"
author: "Daniella Wilkinson"
output: html_document
---

```{r initialize.packages, include=FALSE}

# INSTALL PACKAGES
# install.packages("haven")
# install.packages("dplyr")
# install.packages("table1")
# install.packages("sjPlot")
# install.packages("ggplot2")

# INITIALIZE PACKAGES
library(haven)
library(dplyr)
library(table1)
library(sjPlot)
library(ggplot2)

# ASSIGN WORKING DIRECTORY
getwd()
setwd("~/Desktop/Puerto Rico")
getwd()

```

```{r initialize.datasets, include=FALSE}

# LOAD DATASETS
load("Puerto Rico 2022_puma.RData")
acs <- read_stata("Puerto Rico 2022_acs")
svi <- read.csv("Puerto Rico 2022_svi.csv")

# DATASET 1 
# unique(pumas)
# summary(pumas)

# DATASET 2 
# unique(svi)
# summary(svi)

# DATASET 3
# unique(acs)
# summary(acs)

```

------------------------------------------------------------------------

## Methods

### Data Preparation

#### Data Merging

```{r merge.datasets, include=FALSE}
# PREPARE MERGE : DATASET 1, DATASET 2 by "STCNTY" (State County (Municipality))

# initialize "STCNTY" IN DATASET 1
pumas$STCNTY <- pumas$STATEFP * 1000 + pumas$COUNTYFP

# subset DATASET 2 
subset.DATA1 <- subset(svi, , c(RPL_THEMES, STCNTY, E_TOTPOP))

# EXECUTE MERGE : DATASET 1, DATASET 2 by "STCNTY" (State County (Municipality))
merge.DATA <- merge(pumas, svi, by="STCNTY")

subset.DATA2 <- subset(merge.DATA, , c(RPL_THEMES, STCNTY, E_TOTPOP, PUMA5CE))
# PUMA5CE <- arbitrary municipality numeric ID 
# STCNTY <- literal municipality numeric ID 

# REFINE DATA : Pass 1 

refine.data1 <- aggregate(subset.DATA2, by=list(subset.DATA2$STCNTY), FUN = mean, na.rm=TRUE)

subset.DATA3 <- subset(refine.data1, , c(E_TOTPOP, PUMA5CE))

# REFINE DATA : Pass 2

refine.data2 <- aggregate(subset.DATA3, by=list(refine.data1$PUMA5CE), FUN = sum, na.rm=TRUE)

refine.data2$PUMA5CE <- refine.data2$Group.1
refine.data2$Group.1 <- NULL

refine.data2$PumasPOP <- refine.data2$E_TOTPOP
refine.data2$E_TOTPOP <- NULL

# REFINE DATA : Pass 3 
refine.data3 <- merge(refine.data2, refine.data1, by = "PUMA5CE")

# initialize PUMAS proportion measure
# calculate PUMAS proportion measure
refine.data3$PumasPROP <- refine.data3$E_TOTPOP / refine.data3$PumasPOP
# initialize PUMAS average measure 
# calculate PUMAS average measure
refine.data3$PumasAVG <- refine.data3$PumasPROP * refine.data3$RPL_THEMES

# REFINE DATA : Pass 4 

refine.data4 <- aggregate(refine.data3, by=list(refine.data3$PUMA5CE), FUN=sum)

subset.DATA4 <- subset(refine.data4, , c(Group.1, PumasAVG))

# REFINE DATA : Pass 5 

Puerto.RicoPOP.1 <- merge(subset.DATA4, acs, by.x = "Group.1", by.y = "puma")

# initialize Puerto.RicoPOP.1
# initialize Puerto.RicoPOP.2
Puerto.RicoPOP.2 <- Puerto.RicoPOP.1

```

To analyze the correlational relationship between social vulnerability and socioeconomic factors, we must combine the American Community Survey (ACS), Public Use Microdata Area (PUMA), and Social Vulnerability Index (SVI) datasets into a single respondent-level dataset. This initial step is crucial, enabling us to analyze the social vulnerability experienced by individuals in each Puerto Rican municipality. Therefore, we will perform this multi-step data merging process as described below: 

In our first task, we will extract key variables from the SVI dataset: a respondent's average SVI score '*rpl_themes,'*municipality, '*STCNTY*,' and total population, '*E_TOTPOP*.' This step will allow us to generate a focused subset of the data that contains the relevant information we require from the SVI dataset.

Next, we will merge the PUMA dataset with the SVI dataset using the shared variable '*STCNTY*,' representing the respondent municipalities. This process involves aligning the datasets based on the '*STCNTY*' variable and combining the information from both datasets into a single, unified dataset. With this step completed, we have merged the datasets and may proceed accordingly. The unified dataset displays the corresponding municipality to each respondent in the dataset and directly corresponds to their assigned SVI score.

```{r FILTER.adult.respondent, include=FALSE}

# adult respondents  
Puerto.RicoPOP.2$age <- ifelse(
  Puerto.RicoPOP.1$age >= 18, Puerto.RicoPOP.1$age, NA
)
```

#### Data Re-Coding

The re-coding process is an essential step in the data preparation process. It is imperative to re-code data honestly and logically to undergo our analysis and create additional measures to reinterpret data, combine data to create new relevant measures, and also create semantic labels for variables such as age, race, ethnicity, sex, and gender to increase legibility.

Now that we have cleaned and organized the data, we possess a dataset tailored to the specific parameters we wish to measure. We can now proceed to the next stage of data cleaning, which involves re-coding the necessary variables to ensure both ease and accuracy in our analysis. The specific variables I will be re-coding directly relate to the measures outlined by the Social Vulnerability Index (SVI) and its sub-measures.

For example, I re-coded variables such as age, sex, race, citizenship status, and subsidiary income to improve their clarity. Additionally, I combined marital status and parental status into a new variable called household demographics.

```{r IDENTIFY.SVI.vulnerability}

# Minor Vulnerability
Puerto.RicoPOP.2$PumasAVG.scale [
  Puerto.RicoPOP.2$PumasAVG > 0.00 &
  Puerto.RicoPOP.2$PumasAVG < 0.25
] <- 1

# Moderate Vulnerability
Puerto.RicoPOP.2$PumasAVG.scale [
  Puerto.RicoPOP.2$PumasAVG > 0.25 &
  Puerto.RicoPOP.2$PumasAVG < 0.50
] <- 2

# Significant Vulnerability
Puerto.RicoPOP.2$PumasAVG.scale [
  Puerto.RicoPOP.2$PumasAVG > 0.50 &
  Puerto.RicoPOP.2$PumasAVG < 0.75
] <- 3

# Severe Vulnerability
Puerto.RicoPOP.2$PumasAVG.scale [
  Puerto.RicoPOP.2$PumasAVG > 0.75 &
  Puerto.RicoPOP.2$PumasAVG < 1.00
] <- 4

```

**SVI Interpretation**

The SVI utilizes a quartile system to categorize respondents from most to least vulnerable, rather than relying on arbitrary cutoff points that do not allow for a spectrum of results. Such rigid cutoffs often exclude certain segments of the population from consideration.

I have re-coded the SVI into legible quartiles as suggested by the SVI. This re-coding creates four measures of social vulnerability, categorizing them into minor vulnerability, moderate vulnerability, significant vulnerability, and severe vulnerability. This categorization makes the data analysis portion easier and more legible. If I have a semantic label (a scale measure) for the SVI score, it will provide a clearer understanding of the social vulnerability levels.

```{r RECODE.citizenship.stat, include=FALSE}

# citizenship status 
Puerto.RicoPOP.2$citizenship.stat [
  Puerto.RicoPOP.1$citizen == 0
] <- NA

Puerto.RicoPOP.2$citizenship.stat [
  Puerto.RicoPOP.1$bpl == 110
] <- "Born USA, Citizen"

Puerto.RicoPOP.2$citizenship.stat [
  Puerto.RicoPOP.1$citizen == 1
] <- "Born Abroad, Eligibile Citizen"

Puerto.RicoPOP.2$citizenship.stat [
  Puerto.RicoPOP.1$citizen == 2
] <- "Born Abroad, Naturalized Citizen"

Puerto.RicoPOP.2$citizenship.stat [
  Puerto.RicoPOP.1$citizen == 3 |
  Puerto.RicoPOP.1$citizen == 4
] <- "Residency Status, Other"

Puerto.RicoPOP.2$citizenship.stat [
  Puerto.RicoPOP.1$citizen == 5 |
  Puerto.RicoPOP.1$citizen == 8 |
  Puerto.RicoPOP.1$citizen == 9
] <- "Unkown Citizenship Status"

```

```{r RECODE.socioecon.stat, include=FALSE}

# education attainment 
Puerto.RicoPOP.2$education [
  Puerto.RicoPOP.1$educ == 99
] <- NA

Puerto.RicoPOP.2$education [
  Puerto.RicoPOP.1$educ == 0
] <- "Unknown, No Formal Education"

Puerto.RicoPOP.2$education [
  Puerto.RicoPOP.1$educ == 1
] <- "Elementary"

Puerto.RicoPOP.2$education [
  Puerto.RicoPOP.1$educ == 2
] <- "Junior High"

Puerto.RicoPOP.2$education [
  Puerto.RicoPOP.1$educ == 3 |
  Puerto.RicoPOP.1$educ == 4 |
  Puerto.RicoPOP.1$educ == 5 
] <- "Some High School"

Puerto.RicoPOP.2$education [
  Puerto.RicoPOP.1$educ == 6 
] <- "High School Graduate"

Puerto.RicoPOP.2$education [
  Puerto.RicoPOP.1$educ == 7 |
  Puerto.RicoPOP.1$educ == 8 |
  Puerto.RicoPOP.1$educ == 9 
] <- "Some College"

Puerto.RicoPOP.2$education [
  Puerto.RicoPOP.1$educ == 10 |
  Puerto.RicoPOP.1$educ == 11 
] <- "College Graduate"

# employment status 

Puerto.RicoPOP.2$employment.stat [
  Puerto.RicoPOP.1$empstat == 0 
] <- NA

Puerto.RicoPOP.2$employment.stat [
  Puerto.RicoPOP.1$empstat == 1
] <- "Employed"

Puerto.RicoPOP.2$employment.stat [
  Puerto.RicoPOP.1$empstat == 2 |
  Puerto.RicoPOP.1$empstat == 3
] <- "Unemployed"

Puerto.RicoPOP.2$employment.stat [
  Puerto.RicoPOP.1$empstat == 9 
] <- "Unknown Employment Status"

# receives public subsidiary income 
Puerto.RicoPOP.2$onSUBSIDY <- as.character(
  Puerto.RicoPOP.2$incwelfr)

Puerto.RicoPOP.2$onSUBSIDY <- as.character(
  Puerto.RicoPOP.2$incsupp)

Puerto.RicoPOP.2$onSUBSIDY [
  Puerto.RicoPOP.1$incwelfr == 99999 |
  Puerto.RicoPOP.1$incsupp == 99999 |
  Puerto.RicoPOP.1$foodstmp == 0
] <- "Not Applicable"

Puerto.RicoPOP.2$onSUBSIDY [
  Puerto.RicoPOP.1$incwelfr > 0 |
  Puerto.RicoPOP.1$incsupp > 0 |
  Puerto.RicoPOP.1$foodstmp == 2
] <- 1

Puerto.RicoPOP.2$onSUBSIDY [
  Puerto.RicoPOP.1$incwelfr == 0 |
  Puerto.RicoPOP.1$incsupp == 0 |
  Puerto.RicoPOP.1$foodstmp == 1
] <- 0
```

```{r RECODE.minority.stat, include=FALSE}

# sex characteristics 
Puerto.RicoPOP.2$sex <- as.character(
  Puerto.RicoPOP.1$sex
)

Puerto.RicoPOP.2$sex [
  Puerto.RicoPOP.1$sex == 0
] <- NA

Puerto.RicoPOP.2$sex [
  Puerto.RicoPOP.1$sex == 1
] <- "Male"

Puerto.RicoPOP.2$sex [
  Puerto.RicoPOP.1$sex == 2
] <- "Female"

# race characteristics
Puerto.RicoPOP.2$race <- as.character(
  Puerto.RicoPOP.2$race)

Puerto.RicoPOP.2$race [
  Puerto.RicoPOP.1$race == 1
] <- "White"

Puerto.RicoPOP.2$race [
  Puerto.RicoPOP.1$race == 2
] <- "Black or African American"

Puerto.RicoPOP.2$race [
  Puerto.RicoPOP.1$race == 3
] <- "American Indian or Alaska Native"

Puerto.RicoPOP.2$race [
  Puerto.RicoPOP.1$race == 4 |
  Puerto.RicoPOP.1$race == 5 |
  Puerto.RicoPOP.1$race == 6 
] <- "Asian"

Puerto.RicoPOP.2$race [
  Puerto.RicoPOP.1$race == 7
] <- "Other Race"

Puerto.RicoPOP.2$race [
  Puerto.RicoPOP.1$race == 8 |
  Puerto.RicoPOP.1$race == 9
] <- "Multiracial"

# ethnicity characteristics 
Puerto.RicoPOP.2$hispanic [
  Puerto.RicoPOP.1$hispan == 0
] <- "Not Hispanic"

Puerto.RicoPOP.2$hispanic [
  Puerto.RicoPOP.1$hispan == 1 |
  Puerto.RicoPOP.1$hispan == 2 |
  Puerto.RicoPOP.1$hispan == 3 |
  Puerto.RicoPOP.1$hispan == 4
] <- "Hispanic"

Puerto.RicoPOP.2$hispanic [
  Puerto.RicoPOP.1$hispan == 9
] <- "Unreported"

unique(Puerto.RicoPOP.1$hispan)

# linguistic characteristics
Puerto.RicoPOP.2$eng.isolation [
  Puerto.RicoPOP.1$lingisol == 0
  ] <- NA

Puerto.RicoPOP.2$eng.isolation [
  Puerto.RicoPOP.1$lingisol == 2
] <- 1

Puerto.RicoPOP.2$eng.isolation [
  Puerto.RicoPOP.1$lingisol == 1
] <- 0

# linguistic characteristics
Puerto.RicoPOP.2$eng.proficiency [
  Puerto.RicoPOP.1$speakeng == 0 |
  Puerto.RicoPOP.1$speakeng == 7 |
  Puerto.RicoPOP.1$speakeng == 8 |
  Puerto.RicoPOP.1$speakeng == 9
] <- NA

Puerto.RicoPOP.2$eng.proficiency [
  Puerto.RicoPOP.1$speakeng == 1
] <- "Inproficient"

Puerto.RicoPOP.2$eng.proficiency [
  Puerto.RicoPOP.1$speakeng == 6
] <- "Poor Proficiency"

Puerto.RicoPOP.2$eng.proficiency [
  Puerto.RicoPOP.1$speakeng == 2 |
  Puerto.RicoPOP.1$speakeng == 5 
] <- "Good Proficiency"

Puerto.RicoPOP.2$eng.proficiency [
  Puerto.RicoPOP.1$speakeng == 3 |
  Puerto.RicoPOP.1$speakeng == 4
] <- "Excellent Proficiency"

Puerto.RicoPOP.2$eng.proficiency [
  Puerto.RicoPOP.1$speakeng == 3 
] <- "Monolingual"

# disability characteristics 
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffmob == 0
] <- NA
  
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffmob == 2
] <- 1

Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffmob == 1
] <- 0

# disability characteristics 
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffrem == 0
] <- NA
  
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffrem == 2
] <- 1

Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffrem == 1
] <- 0

# disability characteristics 
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffeye == 0
] <- NA
  
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffeye == 2
] <- 1

Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffeye == 1
] <- 0

# disability characteristics 
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffphys == 0
] <- NA
  
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffphys == 2
] <- 1

Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffphys == 1
] <- 0

# disability characteristics 
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffcare == 0
] <- NA
  
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffcare == 2
] <- 1

Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffcare == 1
] <- 0

# disability characteristics 
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffhear == 0
] <- NA
  
Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffhear == 2
] <- 1

Puerto.RicoPOP.2$disability [
  Puerto.RicoPOP.1$diffhear == 1
] <- 0
```

```{r RECODE.housing.stat, include=FALSE}

# housing payment
Puerto.RicoPOP.2$housing.payment [
  Puerto.RicoPOP.1$ownershp == 0 |
  Puerto.RicoPOP.1$mortgage == 0
] <- "Unreported"

Puerto.RicoPOP.2$housing.payment [
  Puerto.RicoPOP.1$mortgage == 1
] <- "Property Owner"

Puerto.RicoPOP.2$housing.payment [
  Puerto.RicoPOP.1$mortgage == 2 |
  Puerto.RicoPOP.1$mortgage == 3 |
  Puerto.RicoPOP.1$mortgage == 4
] <- "Monthly Mortgage Payment"

Puerto.RicoPOP.2$housing.payment [
  Puerto.RicoPOP.1$ownershp == 2
] <- "Monthly Rental Payment"

# housing payment 
Puerto.RicoPOP.2$add.housing.payment [
  Puerto.RicoPOP.1$mortgag2 == 0 |
  Puerto.RicoPOP.1$mortgag2 == 1
] <- "No Additional Payment"

Puerto.RicoPOP.2$add.housing.payment [
  Puerto.RicoPOP.1$mortgag2 == 2 |
  Puerto.RicoPOP.1$mortgag2 == 3
] <- "2nd Mortgage Payment"

Puerto.RicoPOP.2$add.housing.payment [
  Puerto.RicoPOP.1$mortgag2 == 4
] <- "Home Equity Loan Payment"

Puerto.RicoPOP.2$add.housing.payment [
  Puerto.RicoPOP.1$mortgag2 == 5
] <- "Both"

# housing development 
Puerto.RicoPOP.2$housing.development [
  Puerto.RicoPOP.1$unitsstr == 0
] <- NA

Puerto.RicoPOP.2$housing.development [
  Puerto.RicoPOP.1$unitsstr == 1
] <- "Mobile Home, Trailer Home"

Puerto.RicoPOP.2$housing.development [
  Puerto.RicoPOP.1$unitsstr == 2
] <- "Homeless, Non-traditional Dwelling"

Puerto.RicoPOP.2$housing.development [
  Puerto.RicoPOP.1$unitsstr == 3
] <- "Single-Fam House, Detached"

Puerto.RicoPOP.2$housing.development [
  Puerto.RicoPOP.1$unitsstr == 4
] <- "Single-Fam House, Attached"

Puerto.RicoPOP.2$housing.development [
  Puerto.RicoPOP.1$unitsstr == 5 |
  Puerto.RicoPOP.1$unitsstr == 6 |
  Puerto.RicoPOP.1$unitsstr == 7 |
  Puerto.RicoPOP.1$unitsstr == 8 |
  Puerto.RicoPOP.1$unitsstr == 9 |
  Puerto.RicoPOP.1$unitsstr == 10
] <- "Multi-Fam House, Appt. or Townhouse"

# household demographic
Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 9
] <- NA

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 1 &
  Puerto.RicoPOP.1$nchild > 0 
] <- "Married w/ children, present partner"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 1 &
  Puerto.RicoPOP.1$nchild == 0 
] <- "Married w/o children, present partner"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 2 &
  Puerto.RicoPOP.1$nchild > 0 
] <- "Married w/ children, absent partner"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 2 &
  Puerto.RicoPOP.1$nchild == 0 
] <- "Married w/o children, absent partner"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 3 &
  Puerto.RicoPOP.1$nchild > 0 
] <- "Separeted w/ children"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 3 &
  Puerto.RicoPOP.1$nchild == 0 
] <- "Separated w/o children"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 4 &
  Puerto.RicoPOP.1$nchild > 0 
] <- "Divorced w/ children"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 4 &
  Puerto.RicoPOP.1$nchild == 0 
] <- "Divorced w/o children"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 5 &
  Puerto.RicoPOP.1$nchild > 0 
] <- "Widowed w/ children"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 5 &
  Puerto.RicoPOP.1$nchild == 0 
] <- "Widowed w/o children"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 6 &
  Puerto.RicoPOP.1$nchild > 0 
] <- "Single w/ children"

Puerto.RicoPOP.2$household.dem [
  Puerto.RicoPOP.1$marst == 6 &
  Puerto.RicoPOP.1$nchild == 0 
] <- "Single w/o children"

```

```{r RECODE.trans.stat, include=FALSE}

Puerto.RicoPOP.2$commute [
  Puerto.RicoPOP.1$tranwork == 0
  ] <- "Unreported"

Puerto.RicoPOP.2$commute [
  Puerto.RicoPOP.1$tranwork == 10 |
  Puerto.RicoPOP.1$tranwork == 11 |
  Puerto.RicoPOP.1$tranwork == 12 |
  Puerto.RicoPOP.1$tranwork == 14 |
  Puerto.RicoPOP.1$tranwork == 15 |
  Puerto.RicoPOP.1$tranwork == 20
] <- "Driver, Private vehicle"

Puerto.RicoPOP.2$commute [
  Puerto.RicoPOP.1$tranwork == 13 
] <- "Passenger, Private vehicle"

Puerto.RicoPOP.2$commute [
  Puerto.RicoPOP.1$tranwork == 31 |
  Puerto.RicoPOP.1$tranwork == 32 |
  Puerto.RicoPOP.1$tranwork == 33 |
  Puerto.RicoPOP.1$tranwork == 34 |
  Puerto.RicoPOP.1$tranwork == 35 |
  Puerto.RicoPOP.1$tranwork == 36 |
  Puerto.RicoPOP.1$tranwork == 37 |
  Puerto.RicoPOP.1$tranwork == 38 |
  Puerto.RicoPOP.1$tranwork == 39
] <- "Public Transportation"

Puerto.RicoPOP.2$commute [
  Puerto.RicoPOP.1$tranwork == 50
] <- "On Bike"

Puerto.RicoPOP.2$commute [
  Puerto.RicoPOP.1$tranwork == 60
] <- "On Foot"

Puerto.RicoPOP.2$commute [
  Puerto.RicoPOP.1$tranwork == 70
] <- "Other"

Puerto.RicoPOP.2$commute [
  Puerto.RicoPOP.1$tranwork == 80
] <- "Does Not Commute"

```

#### Data Sub-Setting

Now that the data has been cleaned and sorted, we have a dataset with the exact parameters we are looking to measure. Now we can move on to the next stage of data cleaning which is re-coding the necessary variables to ensure the ease and accuracy of our analysis.

The specific variables I'll be re-coding directly correlate to the measures outlined by the Social Vulnerability Index and its sub-measures which the SVI codes on a quarterly system of most to least vulnerable rather creating arbitrary cutoff measures that do not allow a spectrum of respondent results but rather demands strict adherence to arbitrary cutoffs that often lead a subset of the population out of consideration.

...

Another note, as we prepare our data for subsetting, we need to carefully consider which responses should be labeled as NA and which should be classified as unknown or unreported.

```{r SUBSET.analysis.sample}

Puerto.RicoPOP.3 <- subset(Puerto.RicoPOP.2, ,
  c(
  Group.1, PumasAVG, PumasAVG.scale, age, sex, race, hispanic, disability, citizenship.stat,
  education, eng.proficiency, eng.isolation, employment.stat, onSUBSIDY,
  housing.payment, add.housing.payment, housing.development, household.dem, commute
))

colnames(Puerto.RicoPOP.3)[colnames(Puerto.RicoPOP.3) == "Group.1"] <- "municipality"
colnames(Puerto.RicoPOP.3)[colnames(Puerto.RicoPOP.3) == "PumasAVG"] <- "sviAVG.score"
colnames(Puerto.RicoPOP.3)[colnames(Puerto.RicoPOP.3) == "PumasAVG.scale"] <- "sviAVG.scale"

###

Puerto.RicoPOP.4 <- subset(Puerto.RicoPOP.3, 
    age >= 18 & hispanic == "Hispanic" & citizenship.stat == "Born USA, Citizen" &
    (
      education == "College Graduate" |
      education == "Some College" | 
      education == "High School Graduate" |
      education == "Some High School" 
    ),
  c(
    municipality, sviAVG.score, sviAVG.scale, age, sex, race, hispanic, citizenship.stat,
    education, eng.proficiency, eng.isolation, employment.stat, onSUBSIDY,
    housing.payment, add.housing.payment, housing.development, household.dem
  )
)

analysisPOP <- na.omit(Puerto.RicoPOP.4)
```

### Data Analysis

We have created the final analysis population from the initial sample. By targeting specific variables, we have narrowed our dataset to a manageable subset. With this refined dataset, we can now analyze the data, moving past the data-cleaning stage.

#### Bivariate Analysis

```{r P.Value.Function, include=FALSE}
pvalue <- function(x, ...) {
  # Construct vectors of data y, and groups (strata) g 
  y <- unlist(x) 
  g <- factor(rep(1:length(x), times=sapply(x, length))) 
  if (is.numeric(y)) {
    # Check the number of unique groups 
    num_groups <- length(unique(g)) 
    if (num_groups == 2) {
      p <- t.test(y ~ g)$p.value
    } else { 
      p <- oneway.test(y ~ g)$p.value
    }
  } else {
    p <- chisq.test(table(y, g))$p.value
  }
  # Format the p-value, using an HTML entity for the less-than sign.
  # The initial empty string places the output on the line below the variable label.
  c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}
```

```{r Bivariate.Analysis, echo=FALSE}

analysisPOP$sex <- factor(
  analysisPOP$sex,
  levels = c(
    "Male", "Female"
  ))

analysisPOP$race <- factor(
  analysisPOP$race,
  levels = c(
    "White",
    "Asian",
    "Black or African American",
    "American Indian or Alaska Native",
    "Other Race",
    "Multiracial"
  ))

analysisPOP$education <- factor(
  analysisPOP$education,
  levels = c(
    "College Graduate",
    "Some College",
    "High School Graduate",
    "Some High School",
    "Junior High",
    "Elementary",
    "Unknown, No Formal Education"
  ))

table1(~ age + sex + race + hispanic + citizenship.stat + education + employment.stat + municipality | sviAVG.scale, data = analysisPOP,
       overall = FALSE,
       extra.col = list("P-Value" = pvalue),
       topclass = "Rtable1-zebra",
       caption = "<B>Descriptive Distribution of Population by SVI<B/>",
       footnote = "<i>Caption<i/>"
       )

```

Here I have a table that just displays the most basic demographic data of the people who live in each municipality which includes things like age, race, gender, ethnicity and not much else. I'm just looking at the most basic demographics here. Now, I'll move on to actually testing my hypothesis which includes actually controlling variables like age, race (white, black, multiracial (clear limitation of the dataset, interesting convo about expanding race and ethnicity), hispanc (are hispanic), education (high school and below), and comparing those variables to housing status and transportation.

#### Linear Regression

```{r Linear.Regression}

analysisPOP$employment.stat <- as.factor(analysisPOP$employment.stat)
analysisPOP$education <- factor(analysisPOP$education, 
                                levels = c("Some High School", "High School Graduate", 
                                           "Some College", "College Graduate"))

table(analysisPOP$sviAVG.scale)

m1 <- lm(sviAVG.score ~ employment.stat + education, data = analysisPOP)
m2 <- lm(sviAVG.score ~ education * employment.stat, data = analysisPOP)

summary(m1)
summary(m2)

tab_model(m1)
tab_model(m2)

plot_model( 
  m1, type = "pred", ci.lvl = 0.95,
  terms = c("education", "employment.stat"),
  title = "Predicted Social Vulnerability by Education and Employment"
  )

plot_model(
  m2, type = "pred", ci.lvl = 0.95,
  terms = c("education", "employment.stat"),
  title = "Predicted Social Vulnerability by Education and Employment"
  )
```
